---
title: "Final"
subtitle: "STAT 651"
author: "RJ Cass"
date: "18 Apr. 2026"
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{caption, bm, amsmath, xcolor}
        \setlength{\parindent}{0pt}
        \captionsetup{textfont={it}, font=small} 
        \newcommand{\statvec}[1]{\underset{\sim}{#1}}
geometry: "margin=1in"
---

```{r init}
suppressPackageStartupMessages(library('coda'))
suppressPackageStartupMessages(library('knitr'))
suppressPackageStartupMessages(library('invgamma'))
suppressPackageStartupMessages(library('rstan'))
set.seed(1337)
```


```{r funcs}
run_diagnostics <- function(samples_list) {
  mcmc_obj <- coda::mcmc.list(
    lapply(samples_list, coda::mcmc)
  )

  raftery_results <- lapply(mcmc_obj, coda::raftery.diag)
  names(raftery_results) <- paste0("Chain_", 1:length(raftery_results))

  list(
    Samples = mcmc_obj,
    Gelman = coda::gelman.diag(mcmc_obj),
    Raftery = raftery_results,
    ESS = coda::effectiveSize(mcmc_obj)
  )
}

run_diagnostics_stan <- function(fit_stan) {
  mcmc_obj <- As.mcmc.list(fit_stan)

  raftery_results <- lapply(mcmc_obj, coda::raftery.diag)
  names(raftery_results) <- paste0("Chain_", 1:length(raftery_results))

  list(
    Samples = mcmc_obj,
    Gelman = coda::gelman.diag(mcmc_obj),
    Raftery = raftery_results,
    ESS = coda::effectiveSize(mcmc_obj)
  )
}
```

```{r load_data}
height <- c(10, 14, 17, 21, 24, 28, 31, 35, 39, 42, 46, 49, 53, 56, 60)
distance <- c(174.3, 160.1, 162.8, 174.6, 172.6, 184.4, 181.8, 185.0, 192.4, 193.3, 180.6, 172.5, 181.2, 178.4, 167.6)

n <- length(height)

beta_means_prior <- c(200, 0, 0)
beta_var_prior <- c(75^2, 50^2, 50^2)
error_var_prior_n0 <- 5
error_var_prior_s02 <- 5

error_var_prior_shape <- error_var_prior_n0/2
error_var_prior_scale <- error_var_prior_n0 * error_var_prior_s02/2

height2 <- height*height
```

# 1

### (a)

```{r sampler_1}
X <- model.matrix(distance ~ height + height2)

lm_bayes <- function(y, X, iters = 5000, burnin = 100, 
  inits = list(
    beta = rep(0, ncol(X)),
    gamma = 1/100
  ),
  priors = list(
    # Prior params for betas
    beta_0 = rep(0, ncol(X)),
    Sigma_0_inv = solve(100000*diag(ncol(X))),
    # Prior params for sigma^2
    nu_0 = 0.01,
    sigma2_0 = 1
  )
) {

  n <- nrow(X)
  p <- ncol(X)

  samps <- matrix(nrow = iters, ncol = p+1)

  gamma <- inits$gamma
  beta <- inits$beta

  # Precompute some stuff that will be used in the loop
  XtX <- crossprod(X)
  Xty <- crossprod(X, y)
  Xty_prior <- crossprod(priors$Sigma_0_inv, priors$beta_0)
  shape_prior <- (priors$nu_0 + n)/2
  scale_prior <- priors$nu_0 * priors$sigma2_0

  for (i in 1:(iters+burnin)) {
    # Update beta | gamma, data
    A <- gamma*XtX + priors$Sigma_0_inv
    A_inv <- solve(A)
    B <- gamma*Xty + Xty_prior
    beta <- MASS::mvrnorm(n=1, A_inv%*%B, A_inv)
    
    # Update gamma | beta, data
    SSR <- crossprod(y - X%*%beta)
    gamma <- rgamma(1, shape = shape_prior, scale = .5*(SSR + scale_prior))

    if (i > burnin){
      samps[i-burnin, ] <- c(beta, 1/gamma)
    }
  }
  samps
}

priors <- list(beta_0 = beta_means_prior, Sigma_0_inv = solve(100000*diag(ncol(X))), nu_0 = error_var_prior_n0,
    sigma2_0 = error_var_prior_s02)

# To run on multiple chains
samples <- vector("list", 3)
for (chain in 1:length(samples)) {
  set.seed(1337+chain)
  beta_start <- MASS::mvrnorm(n=1, mu = rep(0, 7), Sigma = 10000*diag(7))
  gamma_start <- rgamma(1, 5, 10)
  samples[[chain]] <- lm_bayes(y = distance, X, inits = list(beta = beta_start, gamma = gamma_start), priors = priors)
}

samples_diag <- run_diagnostics(samples)
all_samps_1 <- as.matrix(samples_diag$Samples)
beta0 <- all_samps_1[, 1]
beta1 <- all_samps_1[, 2]
beta2 <- all_samps_1[, 3]
sigma2 <- all_samps_1[, 4]
post_y <- colMeans(sapply(height, \(x) beta0 + beta1*x + beta2 * x^2 + sigma2))
```

### (b)

The first thing I will explore are the trace plots of the chains:

```{r traceplots_1, fig.height = 5, fig.width = 8}
par(mfrow = c(2, 2))
coda::traceplot(samples_diag$Samples)
par(mfrow = c(1,1))
```

As can be seen from the traceplots, each chain appears to converge to a similar value for each parameter. 

We have the following Effective Sample Sizes for each parameter:

```{r ess_1}
samples_diag$ESS
```

As is seen, we have well over 3000 ESS per parameter. 

Looking at our Gelman values ($\hat{R}$), which compares the variance within a chain to the variance across all chains, we get the following:

```{r gelman_1}
samples_diag$Gelman
```

These values are all fine (a troublesome $\hat{R}$ is usually above 1.1) meaning the variance between chains and within chains is relatively similar (we don't have one chain acting wildly differently fomr the others)

Examining the Raftery values, we have the following:

```{r raftery_1}
samples_diag$Raftery
```

In these tables, the 'Dependence Factor' is the important number, with values around 1 (ie. not high) indicating that we have run our chains long enough to get resonable estimates of parameter values. 

### (c)

The estimates of posterior means and 95% confidence intervals is shown in @tbl-post_est_1.

```{r tbl-post_est_1}
#| tbl-cap: "MonteCarlo estimates of posterior means for each parameter, with 95% equal-tailed intervals"
Estimate <- round(colMeans(all_samps_1), 3)
Lower <- round(apply(all_samps_1, 2, quantile, probs = c(.025)), 3)
Upper <- round(apply(all_samps_1, 2, quantile, probs = c(.975)), 3)
Variable <- c('beta0', 'beta1', 'beta2', 'sigma2')
post_est <- as.data.frame(cbind(Variable, Estimate, Lower, Upper))
rownames(post_est) <- NULL
kable(post_est)
```

### (d)

```{r opt_tee_height}
phi <- -1 * beta1 / (2*beta2)

phi_est <- round(mean(phi), 2)
phi_lower <- round(quantile(phi, probs = c(.025)), 2)
phi_upper <- round(quantile(phi, probs = c(.975)), 2)
```

For the value fo $\phi = \frac{-\beta_1}{2 \beta_2}$ we have a posterior mean of `r phi_est`, with a 95% interval from `r phi_lower` to `r phi_upper`.

### (e)

The distirbution of the data, with the posterior mean estimate of $\phi$ (vertical red line) and the posterior mean estimate of the $y$'s, is provided in @fig-plot_1. As is shown, the posterior estimate of the $y$'s is much higher than the actual distribution, implying the model is at least a decent fit for the data. However, the posterior $\phi$ (the optimum height) does appear to be off a little bit. 

```{r fig-plot_1}
#| fig-cap: "Original data with posterior estimates of Phi and the y's"
plot(height, distance)
abline(h = mean(post_y), col = 'blue')
abline(v = phi_est, col = 'red')
```


# 2

### (a)

We now have $y_i = \alpha - \beta(x_i - \phi)^2 + \epsilon_i$. To show that this is the same as the original likelihood:

$y_i = \alpha - \beta(x_i^2 - 2\phi x_i + \phi^2) + \epsilon_i = \alpha - \beta x_i^2 + 2\beta \phi x_i - \beta \phi^2 + \epsilon_i$

If we let 

$$
\begin{aligned}
\beta_0 &= \alpha - \beta \phi^2  \\
\beta_1 &= 2\beta \phi \\
\beta_2 &= -\beta \\
\epsilon_i &= \epsilon_i
\end{aligned}
$$

We can rewrite this as:

$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \epsilon_i$

### (b)

$p(y|x, \alpha, \beta, \phi, \sigma^2) \sim N(\alpha - \beta(x- \phi)^2, \epsilon) \propto (\epsilon)^{-n/2} e^{-\frac{1}{2\epsilon} \sum(x_i - (\alpha - \beta(x_i - \phi)^2))}$

$p(\alpha) \propto e^{-\frac{1}{2 (50^2)}(\alpha - 220)^2}$

$p(\beta) \propto \beta^{2-1}e^{-10\beta}$

$p(\phi) = e^{-\frac{1}{2 (3^2)}(\phi - 30)^2}$

$p(\sigma^2) = (\sigma^2)^{-5/2 - 1}e^{-\frac{12.5}{(\sigma^2)}}$

$p(y, x, \alpha, \beta, \phi, \sigma^2) = p(y|x, \alpha, \beta, \phi, \sigma^2) p(\alpha) p(\beta) p(\phi) p(\sigma^2)$

$p(\alpha|y, x, \beta, \phi, \sigma^2) \propto p(y|x, \alpha, \beta, \phi, \sigma^2) p(\alpha)$

$p(\beta|y, x, \alpha, \phi, \sigma^2) \propto p(y|x, \alpha, \beta, \phi, \sigma^2) p(\beta)$

$p(\phi|y, x, \alpha, \beta, \sigma^2) \propto p(y|x, \alpha, \beta, \phi, \sigma^2) p(\phi)$

$p(\sigma^2|y, x, \alpha, \beta, \phi) \propto p(y|x, \alpha, \beta, \phi, \sigma^2) p(\sigma^2)$

To skip over the math, I'll just do an MH smapler for EVERY variable. Either that or a PPL. 

### (c)

```{r priors_2}
alpha_prior_mean <- 220
alpha_prior_var <- 50^2
beta_prior_shape <- 2
beta_prior_rate <- 10
phi_prior_mean <- 30
phi_prior_var <- 3^2
sigma2_prior_shape <- 5/2
sigma2_prior_scale <- 5*5/2
```

```{r funcs_2}
loglik_y <- function(X, alpha, beta, phi, epsilon, log_prior) {
  log_lik <- sum(alpha - beta*(X - phi)^2 + epsilon)
  log_lik + log_prior
}

mh_step <- function(X, alpha, beta, phi, epsilon, var, proposal, log_prior_curr, log_prior_prop) {
    log_c <- loglik_y(X, alpha, beta, phi, epsilon, log_prior_curr)
    if (var == 'alpha') {
        curr <- alpha
        log_p <- loglik_y(X, proposal, beta, phi, epsilon, log_prior_prop)
    } else if (var == 'beta') {
        curr <- beta
        log_p <- loglik_y(X, alpha, proposal, phi, epsilon, log_prior_prop)
    } else if (var == 'phi') {
        curr <- phi
        log_p <- loglik_y(X, alpha, beta, proposal, epsilon, log_prior_prop)
    } else {
        curr <- epsilon
        log_p <- loglik_y(X, alpha, beta, phi, proposal, log_prior_prop)
    }

    diff <- log_p - log_c

    if (log(runif(1)) < diff) {
      # Accept
      return(c(res = proposal, acc = 1))
    } else {
        # Reject
        return(c(res = curr, acc = 0))
    }
}
```


```{r mh_sampler}
lm_bayes_2 <- function(y, X, iters = 5000, burnin = 100, 
  inits = list(
    alpha = 0,
    beta = 0,
    phi = 0,
    epsilon = 0
  )
){

  n <- nrow(X)
  p <- ncol(X)

  samps <- matrix(nrow = iters, ncol = 4)

  alpha <- inits$alpha
  beta <- inits$beta
  phi <- inits$phi
  epsilon <- inits$epsilon

  acc_alpha <- 0
  acc_beta <- 0
  acc_phi <- 0
  acc_epsilon <- 0

  for (i in 1:(iters+burnin)) {
    # Update alpha
    alpha_prop <- rnorm(1, mean = alpha_prior_mean, sd = sqrt(alpha_prior_var))
    alpha_log_prior_curr <- dnorm(alpha, mean = alpha_prior_mean, sd = sqrt(alpha_prior_var), log = TRUE)
    alpha_log_prior_prop <- dnorm(alpha_prop, mean = alpha_prior_mean, sd = sqrt(alpha_prior_var), log = TRUE)
    step <- mh_step(X, alpha, beta, phi, epsilon, var = 'alpha', alpha_prop, alpha_log_prior_curr, alpha_log_prior_prop)
    alpha <- step[[1]]
    acc_alpha <- acc_alpha + step[[2]]
    # Update beta
    beta_prop <- rgamma(1, shape = beta_prior_shape, rate = beta_prior_rate)
    beta_log_prior_curr <- dgamma(beta, shape = beta_prior_shape, rate = beta_prior_rate, log = TRUE)
    beta_log_prior_prop <- dgamma(beta_prop, shape = beta_prior_shape, rate = beta_prior_rate, log = TRUE)
    step <- mh_step(X, alpha, beta, phi, epsilon, var = 'beta', beta_prop, beta_log_prior_curr, beta_log_prior_prop)
    beta <- step[[1]]
    acc_beta <- acc_beta + step[[2]]
    # Update phi
    phi_prop <- rnorm(1, mean = phi_prior_mean, sd = sqrt(phi_prior_var))
    phi_log_prior_curr <- dnorm(phi, mean = phi_prior_mean, sd = sqrt(phi_prior_var), log = TRUE)
    phi_log_prior_prop <- dnorm(phi_prop, mean = phi_prior_mean, sd = sqrt(phi_prior_var), log = TRUE)
    step <- mh_step(X, alpha, beta, phi, epsilon, var = 'phi', phi_prop, phi_log_prior_curr, phi_log_prior_prop)
    phi <- step[[1]]
    acc_phi <- acc_phi + step[[2]]
    # Update beta
    epsilon_prop <- rinvgamma(1, shape = sigma2_prior_shape, scale = sigma2_prior_scale)
    epsilon_log_prior_curr <- dinvgamma(epsilon, shape = sigma2_prior_shape, scale = sigma2_prior_scale, log = TRUE)
    epsilon_log_prior_prop <- dinvgamma(epsilon_prop, shape = sigma2_prior_shape, scale = sigma2_prior_scale, log = TRUE)
    step <- mh_step(X, alpha, beta, phi, epsilon, var = 'epsilon', epsilon_prop, epsilon_log_prior_curr, epsilon_log_prior_prop)
    epsilon <- step[[1]]
    acc_epsilon <- acc_epsilon + step[[2]]

    if (i > burnin) {
      samps[i - burnin, ] <- c(alpha, beta, phi, epsilon)
    }
  }

  #cat("Acceptance Rate (alpha):", acc_alpha / iters, "\n")
  #cat("Acceptance Rate (beta):", acc_beta / iters, "\n")
  #cat("Acceptance Rate (phi):", acc_phi / iters, "\n")
  #cat("Acceptance Rate (epsilon):", acc_epsilon / iters, "\n")
  return(samps)
}

```

```{r mh_samples}
samples_2 <- vector("list", 3)
for (s in 1:length(samples_2)) {
    
  alpha_start <- rnorm(1, mean = alpha_prior_mean, sd = sqrt(alpha_prior_var))
  beta_start <- rgamma(1, shape = beta_prior_shape, rate = beta_prior_rate)
  phi_start <- rnorm(1, mean = phi_prior_mean, sd = sqrt(phi_prior_var))
  epsilon_start <- rinvgamma(1, shape = sigma2_prior_shape, scale = sigma2_prior_scale)

  samples_2[[s]] <- lm_bayes_2(y = height, X = distance, iters = 10000, burnin = 100, 
    inits = list(alpha = alpha_start, beta = beta_start, phi = phi_start, epsilon = epsilon_start)
  )
}

samples_diag <- run_diagnostics(samples_2)
all_samps_2 <- as.matrix(samples_diag$Samples)
alpha <- all_samps_2[, 1]
beta <- all_samps_2[, 2]
phi <- all_samps_2[, 3]
epsilon <- all_samps_2[, 4]
post_y <- colMeans(sapply(height, \(x) alpha - beta*(x - phi)^2 + epsilon))
```

```{r}
samples_diag$ESS
```


OK, clearly something is off with my proposals or the likelihood or something, so I will pivot to a PPL. 

```{r stan, message = FALSE, warning = FALSE, output = FALSE}
X <- model.matrix(distance ~ height + height2)

data_stan <- list(N = length(height), y = height, x = X, 
    alpha_mean = alpha_prior_mean, alpha_sd = sqrt(alpha_prior_var), 
    beta_shape = beta_prior_shape, beta_rate = beta_prior_rate, 
    phi_mean = phi_prior_mean, phi_sd = sqrt(phi_prior_var), 
    sig2_shape = sigma2_prior_shape, sig2_scale = sigma2_prior_scale)

params <- c('alpha', 'beta', 'phi', 'epsilon')     

fit_stan <- stan(model_code = readLines("q2.stan"),
                 data = data_stan, pars = params,
                 iter = 10000, warmup = 500, chains = 3)

mcmc_obj <- As.mcmc.list(fit_stan)

samples_diag_2 <- run_diagnostics_stan(fit_stan)
all_samps_2 <- as.matrix(samples_diag_2$Samples)
all_samps_2 <- all_samps_2[, -5]
alpha <- all_samps_2[, 1]
beta <- all_samps_2[, 2]
phi <- all_samps_2[, 3]
epsilon <- all_samps_2[, 4]
post_y <- colMeans(sapply(height, \(x) alpha - beta*(x - phi)^2 + epsilon))
```

STAN Code

```{}
data {
    int<lower=0> N;
    array[N] real y;
    matrix[N,3] x;
    real alpha_mean;
    real alpha_sd;
    real beta_shape;
    real beta_rate;
    real phi_mean;
    real phi_sd;
    real sig2_shape;
    real sig2_scale;
}

parameters {
    real alpha;
    real beta;
    real phi;
    real<lower=0> epsilon;
}

transformed parameters {
  vector[3] betas;
  betas[1] = alpha - beta*phi^2; // Multiply by tau
  betas[2] = 2 * beta * phi;
  betas[3] = -1 * beta;

}

model {
    alpha ~ normal(alpha_mean, alpha_sd);
    beta ~ gamma(beta_shape, beta_rate);
    phi ~ normal(phi_mean, phi_sd);
    epsilon ~ inv_gamma(sig2_shape, sig2_scale);
    
    y ~ normal(x * betas, epsilon);
}
```

### (d)

The first thing I will explore are the trace plots of the chains:

```{r traceplots_2, fig.height = 5, fig.width = 8}
par(mfrow = c(2, 2))
coda::traceplot(samples_diag_2$Samples[, -5])
par(mfrow = c(1,1))
```

As can be seen from the traceplots, each chain appears to converge to a similar value for each parameter. 

We have the following Effective Sample Sizes for each parameter:

```{r ess_2}
samples_diag_2$ESS
```

As is seen, we have well over 3000 ESS per parameter. 

Looking at our Gelman values ($\hat{R}$), which compares the variance within a chain to the variance across all chains, we get the following:

```{r gelman_2}
samples_diag_2$Gelman
```

We have some very poor values here, with values being HUGE. This indicates perhaps poor mixing (requiring perhaps reparameterization, or poor tuning)

Examining the Raftery values, we have the following:

```{r raftery_2}
samples_diag_2$Raftery
```

In this case, some of our chains are ok-sh, but Chain 2 in particular is ROUGH (huge I values), indicating we likely won;t have good prediction from it. 

### (e)

The estimates of posterior means and 95% confidence intervals is shown in @tbl-post_est_1.

```{r tbl-post_est_2}
#| tbl-cap: "MonteCarlo estimates of posterior means for each parameter, with 95% equal-tailed intervals"
Estimate <- round(colMeans(all_samps_2), 4)
Lower <- round(apply(all_samps_2, 2, quantile, probs = c(.025)), 4)
Upper <- round(apply(all_samps_2, 2, quantile, probs = c(.975)), 4)
Variable <- c('alpha', 'beta', 'phi', 'epsilon')
post_est <- as.data.frame(cbind(Variable, Estimate, Lower, Upper))
rownames(post_est) <- NULL
kable(post_est)
```

### (f)


MY POSTERIOR Ys ARE WRONG SO THESE NEXT 2 SECTIONS ARE ALSO JUST WRONG. I didn't have time to figure out what was going wrong. 

For each simulated parameter set, generate a random value for y. Then, do that a bunch of times and find the number that are over 187.

```{r post_prob}

prob_187 <- sapply(35, \(x) alpha - beta*(x - phi)^2 + epsilon)
```

### (g)

Similar to above, but you only do it once to simulate only the next drive. 

```{r next_drive}
# For all simulated parameter values, draw a corresponding prediction for a year
post_preds <- mean(sapply(35, \(x) alpha - beta*(x - phi)^2 + epsilon))
```