---
title: "HW5"
subtitle: "STAT 651"
author: "RJ Cass"
date: "20 Feb. 2026"
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{caption, bm, amsmath, xcolor}
        \setlength{\parindent}{0pt}
        \captionsetup{textfont={it}, font=small} 
        \newcommand{\statvec}[1]{\underset{\sim}{#1}}
geometry: "margin=1in"
---

# 1

Apply a normal approx. to the posterior of the Zero-Inflated Poisson. Plot the contour of the normal approx, with a scatterplot laid on top

```{r}
suppressPackageStartupMessages(library("MASS"))

# Data
y <- c(rep(0, 25), rep(1, 6), rep(2, 4), rep(3, 3), 5)

log_lik <- function(p, lambda, x) {
  # Log-likelihood components
  loglik_zero <- log(p + (1 - p) * exp(-lambda))
  loglik_nonzero <- log(1 - p) - lambda + x * log(lambda) - lgamma(x + 1)
  
  # Combine
  sum(ifelse(x == 0, loglik_zero, loglik_nonzero))
}

logp_unnorm <- function(p, l, y) {
  log_lik(p, l, y) + dbeta(p, 1, 1, log = TRUE) + dgamma(l, 2, 1, log = TRUE)
}

res <- optim(c(0.0, 1.0), \(par) -logp_unnorm(p = par[1], l = par[2], y=y), hessian = TRUE)

appx_mean <- res$par
appx_sigma <- solve(res$hessian)

sim <- mvrnorm(n = 10e3, mu = appx_mean, Sigma = appx_sigma)
dens <- kde2d(x = sim[, 1], y = sim[, 2])
# Plotted below

### Draws from posterior using Gibbs (same code from HW4, for comparison)
logfc_pi <- function(p, lambda, x) {
  log_lik(p, lambda, x) + dbeta(p, 1, 1, log = TRUE)
}

logfc_lambda <- function(p, lambda, x) {
  log_lik(p, lambda, x) + dgamma(lambda, 2, 1, log = TRUE)
}

# Initialize things
n_pi <- 200
pi_seq <- seq(0, 1, length = n_pi)
n_lambda <- 500
lambda_seq <- seq(0, 30, length = n_lambda)

state <- list(pi = .5, lambda = 1.5)
n_iter <- 1000
sims <- as.data.frame(matrix(NA, nrow = n_iter, ncol = length(state)))
colnames(sims) <- names(state)

for (i in 1:n_iter) {
  ## Update Pi
  lp_pi <- sapply(pi_seq, \(p) logfc_pi(p = p, lambda = state$lambda, x = y))
  state$pi <- sample(pi_seq, 1, prob = exp(lp_pi - max(lp_pi))) # It auto-standardizes the probabilities. Max(lp_pi) is to prevent overflow

  ## Update Lambda
  lp_lambda <- sapply(lambda_seq, \(l) logfc_lambda(p = state$pi, lambda = l, x = y))
  state$lambda <- sample(lambda_seq, 1, prob = exp(lp_lambda - max(lp_lambda))) # It auto-standardizes the probabilities. Max(lp_alpha) is to prevent overflow

  sims[i, 'pi'] <- state$pi
  sims[i, 'lambda'] <- state$lambda
}

# Plot the sampled posterior points
plot(sims$pi, sims$lambda, xlab = "pi", ylab = "lambda", main = 'Contour of Norm. Approx. vs. Plot of Sampled Post.') # draws from the joint posterior
contour(dens$x, dens$y, dens$z, col = "red", lwd = 2, add = TRUE)
```

The normal approxiamation appears to be pretty good within the high-density areas: however, I have concerns that it would not capture the 'tail' behaviors of the sampled posterior very well (such as the bottom-left or top-middle of the scatterplot).


# 2

$f(y_i | \eta) = h(y_i)$

Conjugate prior ($n_0, y_0$ are known):

$\pi(\eta) \propto e^{-n_0 A(\eta) + n_0 y_0 \eta}$

#### a) 

$$
\begin{aligned}
\pi(\eta|y)
&\propto L(\eta | y_i)\pi(\eta) \\
&\propto \left[\prod e^{-A(\eta)+y_i \eta} \right] e^{-n_0 A(\eta) + n_0 y_0 \eta} \\
&= e^{-nA(\eta)+ \sum y_i \eta}e^{-n_0 A(\eta) + n_0 y_0 \eta} \\
&= e^{-(n+n_0)A(\eta)+ (n \bar{y} + n_0 y_0) \eta}
\end{aligned}
$$

$\hat{\eta}$ occurs when $\frac{d}{d \eta} \pi_{post} = 0$

For all the following, just assume on all my derivates I have the $|_{\eta = \hat{\eta}}$. I just don't feel like repeating that for every step. 

$$
\begin{aligned}
\frac{d}{d \eta} \pi(\eta|y) 
&= \frac{d}{d \eta}e^{-(n+n_0)A(\eta)+ (n \bar{y} + n_0 y_0) \eta} \\
&= \left[\frac{d}{d \eta} -(n+n_0)A(\eta)+ (n \bar{y} + n_0 y_0) \eta \right] e^{-(n+n_0)A(\eta)+ (n \bar{y} + n_0 y_0) \eta} \\
&= \left[-(n+n_0)[\frac{d}{d \eta} A(\eta)] + (n \bar{y} + n_0 y_0) \right]e^{-(n+n_0)A(\eta)+ (n \bar{y} + n_0 y_0) \eta} 
&= 0
\end{aligned}
$$

The exponential portion can never equal zero, so we need the coefficient to equal 0:

$$
\begin{aligned}
-(n+n_0)[\frac{d}{d \eta} A(\eta)] + (n \bar{y} + n_0 y_0) &= 0 \\
\implies \frac{d}{d \eta} A(\eta) &= \frac{n \bar{y} + n_0 y_0}{n+n_0} \\
\implies \frac{d}{d \eta} A(\eta) &= \frac{n_0}{n_0 + n}y_0 + \frac{n}{n_0 + n}\bar{y}
\end{aligned}
$$

Which matches the required value. 

For the variance, we'll use $\hat{\Sigma} = \left[-\nabla^2 ln(\pi_{post}(\theta|x)) \right]^{-1}$

$$
\begin{aligned}
-\frac{d^2}{d \eta^2} ln(\pi(\eta|y)) 
&= -\frac{d^2}{d \eta^2} \left[ -(n+n_0)A(\eta)+ (n \bar{y} + n_0 y_0) \eta \right] \\
&= -\frac{d}{d \eta} \left[ -(n+n_0)\frac{d}{d \eta}A(\eta) + (n \bar{y} + n_0 y_0) \right] \\
&= (n + n_0) \frac{d^2}{d \eta^2} A(\eta)
\end{aligned}
$$

Plugging this into the above function for $\hat{\Sigma}$:

$\hat{\Sigma} = \left[(n + n_0) \frac{d^2}{d \eta^2} A(\eta) \right]^{-1}$

which is what we needed to show.


#### b) 

$f(y_i | \theta) = Bernoulli(\theta)$. Find a normal approxiamation for posterior of $\eta = ln\left(\frac{\theta}{1-\theta}\right)$

$$
\begin{aligned}
f(y_i | \theta) 
&= \theta^y(1-\theta)^{1-y} = e^{y ln(\theta) + (1-y)ln(1-\theta)} \\
&= e^{y(ln(\theta) - ln(1-\theta)) + ln(1-\theta)} \\
&= e^{y\eta + ln(1 - \frac{e^{\eta}}{1+e^{\eta}})} \\
&= e^{y\eta + ln(\frac{1}{1+e^{\eta}})} \\
&= e^{y\eta - ln(1+e^{\eta})}
\end{aligned}
$$

In this case, $A(\eta) = ln(1+e^{\eta})$

$\frac{d}{d \eta} ln(1+e^{\eta}) = \frac{e^{\eta}}{1+e^{\eta}} = \hat{\theta} = \frac{n_0 y_0 + n \bar{y}}{n_0 + n}$

$$
\begin{aligned}
\frac{d^2}{d \eta^2} ln(1+e^{\eta})
&= \frac{d}{d \eta} e^{\eta}(1+e^{\eta})^{-1} \\
&= e^{\eta}(1+e^{\eta})^{-1} - e^{\eta}e^{\eta}(1+e^{\eta})^{-2} \\
&= \frac{e^{\eta}}{1+e^{\eta}} - \left(\frac{e^{\eta}}{1+e^{\eta}}\right)^2
\end{aligned}
$$

We then have a normal approximation with mean:

$\hat{\theta} = \frac{n_0 y_0 + n \bar{y}}{n_0 + n}$, 

$\hat{\eta} = ln\left(\frac{\hat{\theta}}{1 - \hat{\theta}}\right)$

and variance: 

$\hat{\Sigma} = \left[\frac{e^{\hat{\eta}}(n_0 + n)}{1+e^{\hat{\eta}}} - (n_0 + n)\left(\frac{e^{\hat{\eta}}}{1+e^{\hat{\eta}}}\right)^2 \right]^{-1}$

#### c)

$n_0 = 2, y_0 = 0.5, \sum_{1=1}^8 y_i = 5 \implies n = 8, \bar{y} = \frac{5}{8}$

```{r}
n0 <- 2
y0 <- 0.5
n <- 8
y_bar <- 5/8
sum_y <- 5


theta_hat <- (n0*y0 + n*y_bar)/(n0 + n)
eta_hat <- log(theta_hat/(1-theta_hat))
sigma_hat <- ((n0 + n)*(exp(eta_hat)/(1+exp(eta_hat)) - (exp(eta_hat)/(1+exp(eta_hat)))^2))^(-1)

# Simulate samples
nsim <- 10e3
x <- rnorm(nsim, eta_hat, sqrt(sigma_hat))
w <- exp(-(n0+n)*log(1+exp(x)) + (n0*y0 + sum_y)*x) / dnorm(x, eta_hat, sqrt(sigma_hat))
wn <- w/sum(w)

# P(theta > 0)
p_eta_0 <- sum((x > 0)*wn)
```

Thus, $P(\eta > 0 | y) =$ `r p_eta_0`. 

#### d)

Use a transformation to show the conjugate prior for $\eta$ in the Bernoulli model is same as a Beta prior for $\theta = (1 + e^{-\eta})^{-1}$. Identify shape params as functions of $n_0, y_0$.

$$
\begin{aligned}
\pi(\eta)
&= e^{-n_0 ln(1+e^{\eta}) + n_0y_0\eta} \implies \\
\pi(\theta) &= e^{-n_0 ln(1+e^{ln(\frac{\theta}{1-\theta})}) + n_0y_0ln(\frac{\theta}{1-\theta})}e^{ln(\frac{\theta}{1-\theta})} \\
&= (1+\frac{\theta}{1-\theta})^{-n_0}\left(\frac{\theta}{1-\theta}\right)^{n_0y_0}\frac{\theta}{1-\theta} \\
&= (1-\theta)^{n_0}\theta^{n_0 y_0}(1-\theta)^{-n_0 y_0}\theta(1-\theta)^{-1}\\
&= \theta^{n_0 y_0 + 1}(1-\theta)^{n_0 (1 - y_0) + 1} \\
&= \underline{Beta(n_0 y_0, n_0(1 - y_0))}
\end{aligned}
$$


#### e)

$\pi(\theta|y) \propto \theta^{\sum y_i} (1-\theta)^{n-\sum y_i}\theta^{n_0 y_0}(1-\theta)^{n_0(1-y_0)} = Beta(\sum y_i + n_0 y_0 + 1, n - \sum y_i + n_0(1 - y_0) + 1)$

```{r}
# Transform the previously drawn Etas into Thetas
theta <- 1/(1 + exp(-x))
# Plot densities of sampled thetas and Beta posterior
hist(theta, freq = FALSE, xlim = c(0, 1), ylim = c(0, 3))
curve(dbeta(x, sum_y + n0*y0 + 1, n - sum_y + n0 * (1 - y0) + 1), add = TRUE)
```



# 3

#### a) 

If we take the sum of each row we get: $0 + .4 + .6 = 1, .1 + .3 + .6 = 1, .7 + .2 + .1 = 1$

Thus all rows sum to 1, and this is a valid transition matrix.

#### b) 

To show this is ergodic for MCMC, we need to show that we can get everywhere from anywhere. Basically, we need to show that as n increases, there are no 0s in the transition matrix. 

```{r}
p = matrix(c(0, 0.4, 0.6, 0.1, 0.3, 0.6, 0.7, 0.2, 0.1), nrow = 3, byrow = TRUE)

p%*%p
```

Going just one step in the future we lose the '0' in the transition matrix, so yes, this is a valid transition matrix. 

#### c)

```{r}
p%*%p
```

$P(Y_3 = 3 | Y_1 = 1) = .3$ (the top right cell going from (1) to (3))

#### d)

```{r}
simulate <- function(n_sim, final_step, starting_pos) {
    res <- vector(mode = 'list', length = n_sim)
    for (i in 1:n_sim) {
        pos <- starting_pos
        for (j in 1:(final_step-1)) {
            pos <- sample(c(1, 2, 3), size = 1, prob = p[pos, ])
        }
        res[[i]] <- pos
    }
    res
}

p_y_3 <- mean(simulate(n_sim = 10e3, final_step = 3, starting_pos = 1) == 3)
```

By simluation, $P(Y_3 = 3 | Y_1 = 1) \approx$ `r p_y_3`

#### e) 

```{r}
mult_p <- function(n) {
    p_next <- p
    for (i in 1:(n-1)) {
        p_next <- p_next%*%p
    }
    p_next
}

mult_p(n = 12)
```

Per the matrix, $P(Y_{12} = 1 | Y_1 = 2) = .309$ (left-middle cell: starting at (2), going to (1)).


#### f)

```{r}
p_y_12 <- mean(simulate(n_sim = 10e3, final_step = 12, starting_pos = 2) == 3)
```

By simulation, $P(Y_{12} = 1 | Y_1 = 2) \approx$ `r p_y_12`

#### g)

$P^5$:
```{r}
mult_p(n = 5)
```

$P^{10}$:
```{r}
mult_p(n = 10)
```

$P^{100}$:
```{r}
mult_p(n = 100)
```

#### h)

```{r}
mult_p(n = 10000)
```

The distribution as $n \to \infty$ is $P(1) = .309, P(2) = .2909, P(3) = .4$

#### i)

```{r}
n_sim <- 10000
steps <- vector(mode = 'list', length = n_sim)
steps[[1]] <- 1
for (i in 2:(n_sim)) {
    steps[[i]] <- sample(c(1, 2, 3), size = 1, prob = p[steps[[i-1]], ])
}
p1 <- mean(steps == 1)
p2 <- mean(steps == 2)
p3 <-  mean(steps == 3)
```

By simulation: the longer term probabilities are: $P(1) \approx `r p1`, P(2) \approx `r p2`, P(3) \approx `r p3`$

This appears to be the same as the calculated limit in part (h)

# Acknolwedgements

#### Estimated time

I estimate this assignment took me in total around 5 hours to complete

#### Resources Used

I used some googling to try to understand what Question 2 was asking about in some of the steps (specifically to clarify the different transformations I was supposed to do), as well as class code/slides. 
