---
title: "STAT 651 - Final Project"
author: "Dallin Mason, Andrew Stradling, RJ Cass"
execute:
  echo: false
format: 
  pdf:
    fig-pos: "H"
    include-in-header:
      text: |
        \usepackage{caption, bm, amsmath, xcolor}
        \setlength{\parindent}{0pt}
        \captionsetup{textfont={it}, font=small} 
        \newcommand{\statvec}[1]{\underset{\sim}{#1}}
geometry: "margin=1in"
---

# Introduction

It is generally accepted that certain factors influence your expected income. Some of these factors include race, gender, level of education, etc. We want to use the 1994 US Census data available on Kaggle to determine how certain demographic factors affect a person’s likelihood of earning more than $50k annually. 

```{r load_data, message = FALSE}
suppressPackageStartupMessages(library("rstan"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library('knitr'))
suppressPackageStartupMessages(source('functions.R'))
set.seed(1337)

data <- read_csv("train.csv")

y <- data[,15]

table_dt <- data

data <- data |>
  mutate(
    race = as.factor(race),
    age = as.numeric(scale(age)),
    hours = as.numeric(scale(`hours-per-week`)),
    education = case_when(
      education %in% c("Doctorate", "Masters") ~ "Graduate",
      education %in% c("Bachelors", "Assoc-voc", "Prof-school", "Assoc-acdm", "Some-college") ~ "Undergraduate",
      education %in% c("HS-grad", "12th", "7th-8th", "9th", "10th", "11th") ~ "HighSchool",
      TRUE ~ "Elementary"
    ),
    marital = ifelse(`marital-status` %in% c("Divorced","Separated","Never-married","Widowed"), "Single", "Married")
  )

data$race <- relevel(data$race, 'White')

X <- model.matrix(~ age + hours + education + marital + race + gender, data)[, -1]
X <- scale(X, center = TRUE, scale = FALSE)
y = data$`income_>50K`

colnames(X) <- c('age', 'hours', 'educationGraduate', 'educationHighSchool', 'educationUndergraduate', 'maritalSingle', 'raceAmer-Indian-Eskimo', 'raceAsian-Pac-Islander', 'raceBlack', 'raceOther', 'genderMale')

p <- ncol(X)
```

# Data

The data are extracted from the 1994 United States census. This data was preprocessed and downloaded on the Kaggle website, which obtained the data from the UC Irvine machine learning repository. Each row in the dataset is an individual, where the response (yi) is a binary variable indicating whether the individual’s annual income exceeded $50,000, and a series of demographic factors. Of the available factors, we will use: Age, Hours worked per week, Education Level (highest education achieved), Marital Status, Race, and Gender. The dataset does have other factors available, such as Occupation, but we specifically want to examine the effect of these demographic values not accounting for specific jobs/roles. 

The Education factor has many levels (16), so we decided to bucket these levels into similar groups of: Elementary/Middle, HighSchool, Undergraduate, Graduate. We also condensed MaritalStatus into just whether or not the person is married. Summary statistics of each variable are provided in @tbl-table_continuous and @tbl-table_discrete.

```{r tbl-table_continuous}
#| tabl-cap: Summary of continuous variables
table_dt <- table_dt |>
  mutate(
    education = case_when(
      education %in% c("Doctorate", "Masters") ~ "Graduate",
      education %in% c("Bachelors", "Assoc-voc", "Prof-school", "Assoc-acdm", "Some-college") ~ "Undergraduate",
      education %in% c("HS-grad", "12th", "7th-8th", "9th", "10th", "11th") ~ "HighSchool",
      TRUE ~ "Elementary"
    ),
    marital = ifelse(`marital-status` %in% c("Divorced","Separated","Never-married","Widowed"), "Single", "Married")
  )

table_dt <- table_dt[, c(1, 4, 9, 10, 13, 16)]

mySummary <- function(x) {
  c(mean = round(mean(x[x != 0]), 2), sd = round(sd(x[x != 0]), 2), q25 = round(quantile(x[x != 0], .25), 2), q75 = round(quantile(x[x != 0], .75), 2))
}

reg_summ <- t(apply(table_dt[, c('age', 'hours-per-week')], 2, mySummary))
reg_var_summ <- as.data.frame(cbind(colnames(table_dt[, c('age', 'hours-per-week')]), reg_summ))
colnames(reg_var_summ) <- c('Variable', 'Mean', 'SD', 'Q25', 'Q75')
rownames(reg_var_summ) <- NULL
kable(reg_var_summ)
```

```{r tbl-table_discrete}
#| tabl-cap: Summary of discrete variables
table_dt <- table_dt %>% mutate(across(where(is.character), as.factor))
var_list <- c('education', 'race', 'gender', 'marital')

tot_row <- nrow(table_dt)
table <- data.frame(Variable = character(), Value = character(), Count = character(), stringsAsFactors = FALSE)
for (var in var_list){
  tab <- table(table_dt[[var]])
  tab <- cbind(var, names(tab), tab)
  tab <- tab[tab[,3] != 0, ]
  rownames(tab) <- NULL
  tab <- as.data.frame(tab)
  tab[,3] <- as.numeric(tab[,3])
  tab[, 3] <- paste0(tab[, 3], ' (', round(tab[,3]/tot_row, 3)*100.0, '%)')
  table <- rbind(table, tab)
}

colnames(table) <- c('Variable', 'Value', 'Count (Percent)')
rownames(table) <- NULL
kable(table)
```

# Model

We plan to fit a logistic regression model, described as follows: 

$$
\begin{aligned}
y_i \sim & Bernoulli(p_i) \\ 
logit(p_i) = η_i = & β_0 + β_1 Age + β_2 HoursPerWeek + \sum_{j=1}^3 β_{3,j} Education_j + β_4 MaritalStatus \\
& + \sum_{k=1}^4 β_{5,k}Race_k + β_6 GenderMale
\end{aligned}
$$

In this model, the Education, MaritalStatus, Race, and Gender variables are indicator variables which indicate whether or not a certain person falls into that category. Age and HoursPerWeek are continuous variables so the corresponding $\beta$’s represent the slope of the effect of those factors. 


```{r load_priors}
alpha_var <- 10

beta_means <- rep(0, p) # Betas default to 0 mean (updated later for age and hours)
beta_sds   <- rep(0.7, p) # Betas default to .7 var (updated later for age and hours)

# optional informative priors
beta_means[grep("^age", colnames(X))]   <- 0.05
beta_means[grep("^hours", colnames(X))] <- 0.03

beta_sds[grep("^age", colnames(X))]   <- 0.5
beta_sds[grep("^hours", colnames(X))] <- 0.5

# The non-white Race have hierarchical prior
super_p <- 1 # Number of shared parameters
shared_inds <- rep(0, p)
race_cols <- grep("^race", colnames(X))
shared_inds[race_cols] <- 1

# Hierarchical variance for race
race_shared_mean <- 0
race_var <- 0.5
race_mean_var <- 1
```

# Priors

For each set of $\beta$'s we picked a normal prior. For the continuous variables (Age and HoursPerWeek), because we expect these $\beta$'s to be positive (we expect older people to have a higher likelihood of earning more than \$50k), we picked a slightly-positive mean with a smaller standard deviation so it is more weighted towards positive values. For the other $\beta$'s we picked the mean to be 0, with wider standard deviations as less-informative priors. The specific values used are provided in @tbl-priors.

|Variable| $\beta$ | Prior | Chosen Values |
|:---:|:--:|:--------:|:--------------------:|
|Intercept| 0 | $N(0, \sigma^2)$ | $\sigma^2 = `r alpha_var`$|
|Age| 1 | $N(\pi, \gamma^2)$ | $\pi = `r beta_means[1]`, \gamma = `r beta_sds[1]`$|
|HoursPerWeek| 2 | $N(\kappa, \psi^2)$ | $\kappa = `r beta_means[2]`, \psi = `r beta_sds[2]`$|
|Education| 3-5 | $N(\theta, \nu^2)$ | $\theta = `r beta_means[3]`, \nu = `r beta_sds[3]`$|
|MaritalStatus| 6 | $N(\mu, \tau^2)$ | $\mu = `r beta_means[6]`, \tau = `r beta_sds[6]`$|
|Race| 7-10 | $N(\lambda, \rho^2)$ | $\lambda = `r beta_means[7]`, \rho = `r beta_sds[7]`$|
|Gender| 11 | $N(\phi, \epsilon^2)$ | $\phi = `r beta_means[11]`, \epsilon = `r beta_sds[11]`$|
: Chosen priors for the analysis {#tbl-priors}

We also implemented a hierarchical prior on some of the race categories, where we said the means of the non-white race categories have a shared $N(\lambda_s, \rho_s^2), \lambda_s = `r race_shared_mean`, \rho_s = `r race_mean_var`$ prior. This will enable them to 'share strength' as we believe these race categories should behave similarly. 

# Sampler Performance

```{r fake_true_data}
p <- 11
super_p <- 1
alpha_true <- -2

shared_inds <- rep(0, p)
# Race - Nonwhite
shared_inds[7:10]   <- 1

n <- 100

fake_X <- generate_X(n)

beta_true <- generate_beta_true(p, shared_inds, alpha_var, prior_means = beta_means, prior_sds = beta_sds, race_hier_mean = race_shared_mean, race_mean_var)

fake_y <- generate_data(n, X, beta_true)
```

```{r ppl_coverage_sim, message = FALSE, warning = FALSE}
ppl_iter <- 2000
ppl_burn_in <- 500
data_stan <- generate_stan_data(n, p, fake_X, super_p, fake_y, shared_inds, alpha_var, prior_means = beta_means, prior_sds = beta_sds, race_hier_mean = race_shared_mean, race_mean_var, race_cols)

n_cores <- parallel::detectCores()
options(mc.cores = n_cores)          

fit_stan <- stan(
  model_code = readLines("final_vec.stan"),
  data = data_stan,
  pars = c("beta"),
  iter = ppl_iter,
  warmup = ppl_burn_in,
  chains = n_cores
)
post <- rstan::extract(fit_stan)

ppl_beta_est   <- apply(post$beta, 2, mean)
ppl_beta_lower <- apply(post$beta, 2, quantile, 0.025)
ppl_beta_upper <- apply(post$beta, 2, quantile, 0.975)
```

```{r manual_coverage_sim}
mcmc_iter <- 2000
mcmc_burn_in <- 500

out <- run_mcmc(X = fake_X, y = fake_y, alpha_var = alpha_var, beta_means = beta_means, beta_sds = beta_sds,
race_var = race_var, race_mean_var = race_mean_var, shared_inds = shared_inds, n_iter = mcmc_iter)

samples <- out$samples

mcmc_beta_est   <- apply(samples, 2, mean)
mcmc_beta_lower <- apply(samples, 2, quantile, 0.025)
mcmc_beta_upper <- apply(samples, 2, quantile, 0.975)
```

To measure our model performance, we generated a set of fake data using 'true' parameters based on our priors. We ran each sampler on this fake data to see if each sampler could capture the true values. As shown in @fig-coverage_plots, both samplers are able to capture almost all of the true parameters within their posterior 95% confidence intervals. This gives us confidence that the samplers are performing the sampling expected and correctly exploring the required space for each parameter. 

```{r fig-coverage_plots, fig.width = 7, fig.height = 4}
#| fig-cap: "Coverage of each sampler. Blue Triangles: true value. Black Cicle: Posterior Mean. Bars: 95% interval of the posterior estimate."
par(mfrow = c(1, 2))
ord <- c(1:2,3:6,7:10,11)
ypos <- p:1  

# PPL
plot(ppl_beta_est[ord], ypos, xlim = range(c(ppl_beta_lower, ppl_beta_upper)), pch = 19,
     yaxt = "n", ylab = "Coefficient", xlab = "Beta value",
     main= paste0('PPL: n - ', n, ', n_sim - ', ppl_iter))

segments(ppl_beta_lower[ord], ypos, ppl_beta_upper[ord], ypos)

axis(2, at = ypos, labels = ord, las = 2)
points(beta_true[ord], ypos, col = "blue", pch = 17)

# MCMC
plot(mcmc_beta_est[ord], ypos, xlim = range(c(mcmc_beta_lower, mcmc_beta_upper)), pch = 19,
     yaxt = "n", ylab = "Coefficient", xlab = "Beta value",
     main= paste0('MCMC: n - ', n, ', n_sim - ', mcmc_iter))

segments(mcmc_beta_lower[ord], ypos, mcmc_beta_upper[ord], ypos)

axis(2, at = ypos, labels = ord, las = 2)
points(beta_true[ord], ypos, col = "blue", pch = 17)

par(mfrow = c(1, 1))
```

# Model Fitting

```{r mcmc}
mcmc_iter <- 2000
mcmc_burn_in <- 500

mcmc_start_time <- Sys.time()
out <- run_mcmc(X = X, y = y, alpha_var = alpha_var, beta_means = beta_means, beta_sds = beta_sds,
race_var = race_var, race_mean_var = race_mean_var, shared_inds = shared_inds, n_iter = mcmc_iter)
mcmc_end_time <- Sys.time()
mcmc_run_time <- round(as.numeric(mcmc_end_time - mcmc_start_time, units = 'secs'))

# Checking acceptance rates
# out$accept_alpha
# out$accept_beta
# out$accept_shared

samples <- out$samples

ess <- coda::effectiveSize(samples)
#discard burnin
samples <- samples[-c(1:mcmc_burn_in),]

# Posterior means
beta_hat <- colMeans(samples)

# Posterior SDs
beta_sd <- apply(samples, 2, sd)
```

```{r ppl, message = FALSE, warning = FALSE, include = FALSE}
ppl_iter <- 2000
ppl_burn_in <- 500
data_stan <- list(n = nrow(data), p = p, X = X, super_p = super_p, shared_inds = shared_inds, y = y,
                  alpha_var = alpha_var, prior_means = beta_means, prior_sds = beta_sds, 
                  race_hier_mean = race_shared_mean, race_hier_var = race_mean_var, race_cols = race_cols)

params <- c("alpha", "beta", "shared_mean")

n_cores <- parallel::detectCores()
options(mc.cores = n_cores)          

ppl_start_time <- Sys.time()
fit_stan <- stan(model_code = readLines("final_vec.stan"), data = data_stan, pars = params,
                 iter = ppl_iter, warmup = ppl_burn_in, chains = 1)
ppl_end_time <- Sys.time()
ppl_run_time <- round(as.numeric(ppl_end_time - ppl_start_time, units = 'secs'))

sim_stan <- rstan::extract(fit_stan) # collect samples from all chains
```

We fit the model using a Metropolis-Hastings sampler, as well as a sampler in STAN using Hamiltonian Monte Carlo. The structure of our model does not allow for any conjugate priors, meaning every update step in our sampler was necessarily done using Metropolis-Hastings. The M-H sampler performed `r mcmc_iter` in `r mcmc_run_time` seconds while the PPL sampler performed `r ppl_iter` in `r ppl_run_time` seconds. The PPL took significantly longer than the MCMC sampler, even after extensive efforts to optimize the sampler performance. A comparison of the posterior distributions of each parameter are provided in @fig-post_betas.

In both samplers, the posterior distributions for the continuous variables, and the factors with only 2 levesl (Gender and MaritalStatus), have comparatively tight posterior distribution. The factors with multiple levels (Education and Race) have much wider distributions. However, both samplers have resulted in very similar posterior distributions for each parameters, although the MCMC sampler tended to have slightly lower parameter estimates. 

```{r fig-post_betas, fig.width = 7, fig.height = 4}
#| fig-cap: "Posterior distributions of each parameter. The continuous variables, and the factors with only 2 levels, are well estimated. The factors with multiple levels have much wider distirbutions."

beta_list <- c("Age", "HoursPerWeek", "Edu - Grad", "Edu - H.S.", "Edu - U.Grad", "Mar. - Single", "Race - Amer-Indian", "Race - Asian", "Race - Black", "Race - Other", "Gender - Male")

ppl_long <- as.data.frame(sim_stan$beta) %>% pivot_longer(cols = everything(), names_to = "Parameter", values_to = "Value") %>% mutate(Source = "PPL")
mcmc_long <- as.data.frame(samples) %>% pivot_longer(cols = everything(), names_to = "Parameter", values_to = "Value") %>% mutate(Source = "MCMC")

# Combine datasets
combined_df <- bind_rows(ppl_long, mcmc_long)

combined_df$Parameter <- rep(beta_list, 2*(ppl_iter-ppl_burn_in))

combined_df <- combined_df %>%
  mutate(Parameter = factor(Parameter, 
                            levels = beta_list))

# Create the paired boxplot
ggplot(combined_df, aes(x = Parameter, y = Value, fill = Source)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Posterior Distributions - PPL vs. MCMC", x = "Parameters", y = "Posterior Distribution") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.position = "inside", legend.position.inside = c(.93, .85))
```


As mentioned previously, we also placed a hierarchical prior on the non-white categories of race. This is because we feel these categories should perform somewhat similarly and should have a similar population mean when compared to Race - White. To visualize the 'sharing of strength' between these parameters, we fit a non-bayesian linear regression to get regular parameter estimates, and compared that to the parameters estimates from our Bayesian sampling. As shown in @fig-stan_sharing_of_strength, the individual parameter estimates from the PPL are slightly 'shrunk' towards the posterior shared mean (primarily identifiable in American-Indian-Eskimo).

```{r normal_regression, message = FALSE, warning = FALSE, include = FALSE}
unpooled_fit <- glm(y ~ age + hours + education + marital + race + gender, 
                    data = data, family = "binomial")
unpooled_betas <- coef(unpooled_fit)[race_cols+1]
reg_betas <- coef(unpooled_fit)[-1]
reg_betas_ci <- confint(unpooled_fit)[-1, ]
```

```{r manual_sharing_of_strength}
# # Extract your hierarchical posterior means from the MCMC 'samples'
# hier_betas <- colMeans(samples[, race_cols])

# # Combine into a data frame
# plot_data <- data.frame(
#   Race = colnames(X)[race_cols],
#   Unpooled = unpooled_betas,
#   Hierarchical = hier_betas
# )

# ggplot(plot_data) +
#   # Draw a vertical line at the shared mean to show the 'target' of shrinkage
#   geom_vline(xintercept = mean(sim_stan$shared_mean), 
#              linetype = "dashed", color = "gray50") +
#   # Draw the connecting lines
#   geom_segment(aes(x = Unpooled, xend = Hierarchical, y = Race, yend = Race),
#                arrow = arrow(length = unit(0.2, "cm")), color = "blue") +
#   # Plot the two types of points
#   geom_point(aes(x = Unpooled, y = Race, color = "Unpooled (No Sharing)"), size = 3) +
#   geom_point(aes(x = Hierarchical, y = Race, color = "Hierarchical (Borrowing Strength)"), size = 3) +
#   labs(title = "MCMC: Shrinkage in Race Coefficients",
#        subtitle = "Arrows show how hierarchical priors pull group estimates toward the common mean",
#        x = "Coefficient Value (Log-Odds)",
#        y = "Race Category",
#        color = "Estimation Method") +
#   theme_minimal()

```

```{r fig-stan_sharing_of_strength}
#| fig-cap: "Representation of the sharing of strength of the non-white race parameters. Notice that the red dots (the estimates from the PPL) are closer to the 'true value' when compared to the estimates from a non-bayesian linear regression. The MCMC sampler had almost identical results."

# Extract your hierarchical posterior means from the MCMC 'samples'
hier_betas <- colMeans(sim_stan$beta[, race_cols])

# Combine into a data frame
plot_data <- data.frame(
  Race = colnames(X)[race_cols],
  Unpooled = unpooled_betas,
  Hierarchical = hier_betas
)

ggplot(plot_data) +
  # Draw a vertical line at the shared mean to show the 'target' of shrinkage
  geom_vline(xintercept = mean(sim_stan$shared_mean), 
             linetype = "dashed", color = "gray50") +
  # Draw the connecting lines
  geom_segment(aes(x = Unpooled, xend = Hierarchical, y = Race, yend = Race),
               arrow = arrow(length = unit(0.2, "cm")), color = "blue") +
  # Draw the connecting lines
  geom_segment(aes(x = Unpooled, xend = Hierarchical, y = Race, yend = Race),
               arrow = arrow(length = unit(0.2, "cm")), color = "blue") +
  # Plot the two types of points
  geom_point(aes(x = Unpooled, y = Race, color = "Unpooled (No Sharing)"), size = 3) +
  geom_point(aes(x = Hierarchical, y = Race, color = "Hierarchical (Borrowing Strength)"), size = 3) +
  labs(title = "PPL: Shrinkage in Race Coefficients",
       subtitle = "Arrows show how hierarchical priors pull group estimates toward the common mean",
       x = "Coefficient Value (Log-Odds)",
       y = "Race Category",
       color = "Estimation Method") +
  theme_minimal()

```


# Comparison to Non-Bayesian Analysis
To further compare the performance of the non-bayesian regression versus our Bayesian samplers, we compared the distributions of the parameter estimates. As shown in @fig-ppl_vs_reg, both methods provide very similar estimates. However, many of the parameter estimates are slightly lower in the PPL method as compared to the linear regression. The box-plots show the full posterior distribution: however, the 95% confidence interval on the mean of the posteriors (not shown) are much smaller than the 95% confidence interval on the mean from linear regression. 

```{r fig-ppl_vs_reg, fig.width = 7, fig.height = 4}
#| fig-cap: "Posterior distirbutions of the parameters vs. the 95% confidence intervals on the mean from non-bayseian linear regression."

beta_list <- c("Age", "HoursPerWeek", "Edu - Grad", "Edu - H.S.", "Edu - U.Grad", "Mar. - Single", "Race - Amer-Indian", "Race - Asian", "Race - Black", "Race - Other", "Gender - Male")

ppl_long <- as.data.frame(sim_stan$beta) %>% pivot_longer(cols = everything(), names_to = "Parameter", values_to = "Value") %>% mutate(Source = "PPL")

ppl_long$Parameter <- rep(beta_list, (ppl_iter-ppl_burn_in))

summary_data <- data.frame(
  Parameter = beta_list,
  mean = reg_betas,
  lower = reg_betas_ci[, 1],
  upper = reg_betas_ci[, 2]
)

ggplot(ppl_long, aes(x = Parameter, y = Value)) +
  # 1. Map 'fill' to a label inside aes()
  geom_boxplot(aes(fill = "Posterior Distribution"), alpha = 0.5) +
  # 2. Map 'color' to a label inside aes()
  geom_errorbar(data = summary_data, 
                aes(y = mean, ymin = lower, ymax = upper, color = "Regression Estimate"),
                width = 0.2, linewidth = 1,
                position = position_nudge(x = 0.25)) + 
  geom_point(data = summary_data, 
             aes(y = mean, color = "Regression Estimate"), 
             size = 2,
             position = position_nudge(x = 0.25)) +
  # 3. Define the colors manually
  scale_fill_manual(name = NULL, values = c("Posterior Distribution" = "lightblue")) +
  scale_color_manual(name = NULL, values = c("Regression Estimate" = "red")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = 'inside',
    legend.position.inside = c(0.95, 0.95),
    legend.justification = c("right", "top"),
    legend.background = element_rect(fill = "white", color = "grey80")
  ) +
  labs(title = "PPL Posterior and Regression 95% Confidence Intervals", y = "Value", x = "Parameters")
```

A few clear conclusions stand out from this plot. First, older individuals and those who work more hours in a week are significantly more likely to earn an annual salary greater than $50k. We also notice that obtaining higher levels of education increases the likelihood of high earnings. Overall, according to our model, the individuals who are most likely to earn salaries in excess of $50k are older, white, male, have a graduate degree, and are married. Though most of these conclusions are somewhat unsurprising, it is valuable to have more concrete evidence for these phenomena.

# Sensitivity to Priors

# Alternative Prior Analysis

```{r change priors}
alpha_var_mod <- 2   # tighter intercept (was 10)

beta_means_mod <- rep(0, p)

# strong shrinkage toward 0
beta_sds_mod <- rep(0.3, p)

# keep slight directional belief but weaken it
beta_means_mod[grep("^age", colnames(X))]   <- 0.02
beta_means_mod[grep("^hours", colnames(X))] <- 0.01

beta_sds_mod[grep("^age", colnames(X))]   <- 0.3
beta_sds_mod[grep("^hours", colnames(X))] <- 0.3

# hierarchical structure stays (important—you're isolating shrinkage, not removing structure)
super_p_mod <- 1
shared_inds_mod <- rep(0, p)

race_cols_mod <- grep("^race", colnames(X))
shared_inds_mod[race_cols_mod] <- 1

# stronger pooling across race categories
race_shared_mean_mod <- 0
race_var_mod <- 0.3
race_mean_var_mod <- 0.3
```



As a sensitivity analysis, we considered an alternative prior specification that imposes stronger shrinkage on the regression coefficients. In this setting, we reduced the prior variances of the coefficients from their original values (e.g., standard deviations of 0.7 and 0.5) to a smaller common value of 0.3, and similarly reduced the intercept variance from 10 to 2. This reflects a more conservative prior belief that most predictors have relatively small effects on the response and should be shrunk toward zero unless strongly supported by the data. For the continuous predictors Age and HoursPerWeek, we retained a slight positive prior mean to reflect expected directionality, but reduced both the magnitude of the mean and the variance to limit their influence. Additionally, we strengthened the hierarchical prior on race effects by reducing the associated variance components, thereby increasing the degree of partial pooling across categories. This alternative specification allows us to assess whether our substantive conclusions are robust to a more strongly regularizing and skeptical prior.The specific values used are provided in @tbl-priors-alt.

|Variable| $\beta$ | Prior | Chosen Values |
|:----:|:--:|:--------:|:--------------------:|
|Intercept| 0 | $N(0, \sigma^2)$ | $\sigma^2 = `r alpha_var_mod`$|
|Age| 1 | $N(\pi, \gamma^2)$ | $\pi = `r beta_means_mod[1]`, \gamma = `r beta_sds_mod[1]`$|
|HoursPerWeek| 2 | $N(\kappa, \psi^2)$ | $\kappa = `r beta_means_mod[2]`, \psi = `r beta_sds_mod[2]`$|
|Education| 3-5 | $N(\theta, \nu^2)$ | $\theta = `r beta_means_mod[3]`, \nu = `r beta_sds_mod[3]`$|
|MaritalStatus| 6 | $N(\mu, \tau^2)$ | $\mu = `r beta_means_mod[6]`, \tau = `r beta_sds_mod[6]`$|
|Race| 7-10 | $N(\lambda, \rho^2)$ | $\lambda = `r beta_means_mod[7]`, \rho = `r beta_sds_mod[7]`$|
|Gender| 11 | $N(\phi, \epsilon^2)$ | $\phi = `r beta_means_mod[11]`, \epsilon = `r beta_sds_mod[11]`$|
: Table of alternate priors for sensitivity analysis {#tbl-priors-alt}

```{r alt_priors_ppl}
ppl_iter_mod <- 2000
ppl_burn_in_mod <- 500
data_stan_mod <- generate_stan_data(n, p, fake_X, super_p, fake_y, shared_inds, alpha_var, prior_means = beta_means_mod, prior_sds = beta_sds_mod, race_hier_mean = race_shared_mean_mod, race_mean_var_mod, race_cols_mod) 

fit_stan_mod <- stan(
  model_code = readLines("final_vec.stan"),
  data = data_stan_mod,
  pars = c("beta"),
  iter = ppl_iter_mod,
  warmup = ppl_burn_in_mod,
  chains = n_cores
)
post_mod <- rstan::extract(fit_stan_mod)

ppl_beta_est_mod   <- apply(post_mod$beta, 2, mean)
ppl_beta_lower_mod <- apply(post_mod$beta, 2, quantile, 0.025)
ppl_beta_upper_mod <- apply(post_mod$beta, 2, quantile, 0.975)
```

```{r fig-coverage_plots_alt, fig.width = 7, fig.height = 4}
#| fig-cap: "Coverage of each PPL - Adjusted vs Not. Blue Triangles: true value. Black Cicle: Posterior Mean. Bars: 95% interval of the posterior estimate."
par(mfrow = c(1, 2))
ord <- c(1:2,3:6,7:10,11)
ypos <- p:1 

# PPL Non-adjusted prior
plot(ppl_beta_est[ord], ypos, xlim = range(c(ppl_beta_lower, ppl_beta_upper)), pch = 19,
     yaxt = "n", ylab = "Coefficient", xlab = "Beta value",
     main= paste0('PPL Original Prior'))

segments(ppl_beta_lower[ord], ypos, ppl_beta_upper[ord], ypos)

axis(2, at = ypos, labels = ord, las = 2)
points(beta_true[ord], ypos, col = "blue", pch = 17)

# PPL Adjusted Prior
plot(ppl_beta_est_mod[ord], ypos, xlim = range(c(ppl_beta_lower_mod, ppl_beta_upper_mod)), pch = 19,
     yaxt = "n", ylab = "Coefficient", xlab = "Beta value",
     main= paste0('PPL Adjusted Prior'))

segments(ppl_beta_lower_mod[ord], ypos, ppl_beta_upper_mod[ord], ypos)

axis(2, at = ypos, labels = ord, las = 2)
points(beta_true[ord], ypos, col = "blue", pch = 17)

par(mfrow = c(1, 1))
```

Under the stronger shrinkage prior, we observe that a subset of coefficients (notably $\beta_3$, $\beta_4$, $\beta_5$, and $\beta_{10}$) remained relatively stable, with posterior means similar to those obtained under the original prior and with slightly reduced standard deviations. This suggests that these effects are strongly supported by the data and are robust to the choice of prior. In contrast, the remaining coefficients were more noticeably affected by the increased regularization, with their posterior means shrinking toward zero and their associated uncertainty increasing. This indicates that these effects are weaker and more sensitive to prior assumptions. Overall, the results suggest that only a subset of predictors have robust, well-identified effects, while others depend more heavily on the prior specification.

# Summary

We used the 1994 US Census data to fir a model of the likelihood that an individual earns more than $50k. We fit a model using a manual Metropolis-Hastings sampler, a PPL (STAN), and off-the-shelf GLM. We generated a fake dataset using 'true' parameters and examined the coverage of the samplers. We identified that both samplers have reasonable coverage (the PPL captured all the true betas, and the M-H missed only one). We then used both samplers on the real dataset and compared the posterior distributions for each beta. We found the distributions to be very similar, but with the PPL yielding slightly higher estimates for the mean, and slightly wider distributions.

Comparing the samplers to the GLM, we found that all the models performed relatively the same. However, the GLM has much higher Education beta estimates than either of our samplers. It also resulted in larger confidence intervals on the means of the estimates as compared to our samples. We also compared the GLM estimates for the race parameters (on which we had a hierarchical prior) to the estimates from the PPL and were able to identify a small portion of sharing of strength.

We identified that those most likely to earn salaries greater than $50k are older, married white males who work longer hours and have a graduate degree. Though somewhat unsurprising results, both samplers and the GLM agree on these effects. These inferences are valuable to understanding an individual's potential for higher earnings in the future. 

Finally, we conducted a small study to identify sensitivity to the priors. We identified that most parameters were relatively robust to the prior values. However, some of the race parameters changed a decent amount, with one even going outside of the coverage bounds.


{{< pagebreak >}}

# Appendix

### Code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

```{r}
#| file: "functions.R"
#| echo: true
```


### GenAI Disclosure

We used GenAI to assist in troubleshooting our PPL and manual sampler. The structure of the code was written by us (ie. the function outlines, step processes, etc.), but we needed a lot of help from chat in terms of optimizing the flows. Chat showed us how to code the standard-normal parameterizations in the manual sampler and PPL. We also used chat to get the basic structure around how to generate some of the graphs (mainly the multi-boxplot charts). 