---
title: "HW8"
subtitle: "STAT 651"
author: "RJ Cass"
date: "3 Apr. 2026"
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{caption, bm, amsmath, xcolor}
        \setlength{\parindent}{0pt}
        \captionsetup{textfont={it}, font=small} 
        \newcommand{\statvec}[1]{\underset{\sim}{#1}}
geometry: "margin=1in"
---


# 1 

With the same fit (Hoff 8.3) from HW7, do the following:

### a)

Perform posterior check with $T(y, \theta) = \underset{j}{max}(Var(y_{i,j})) - \underset{j}{min}(Var(y_{i,j}))$

STAN code for parts (a)-(e)
```{}
data {
    int<lower=0> N; //Number of Schools
    int<lower=0> n; //Total datapoints
    array[n] real y; // Responses
    array[n] int grp; //Associated group for each response
    real nu_0;
    real sigma2_0;
    real eta_0;
    real tau2_0;
    real mu_0;
    real gamma2_0;
}

parameters {
    array[N] real theta;
    real mu;
    real<lower=0> sigma2;
    real<lower=0> tau2;
    
}

transformed parameters {
    array[n] real theta_vec;
    for (i in 1:n) {
        theta_vec[i] = theta[grp[i]];
    }
}

model {
    y ~ normal(theta_vec, sqrt(sigma2));
    theta ~ normal(mu, sqrt(tau2));
    sigma2 ~ inv_gamma(nu_0/2, nu_0 * sigma2_0/2);
    tau2 ~ inv_gamma(eta_0/2, eta_0 * tau2_0/2);
    mu ~ normal(mu_0, sqrt(gamma2_0));
}

generated quantities {
    vector[n] y_rep;
    vector[n] y_upc; 
    vector[n] y_llik; 
    vector[N] theta_upc; 

    for (i in 1:n) { 
        real theta_i = theta[grp[i]]; 
        y_rep[i] = normal_rng(theta_i, sqrt(sigma2)); 
        y_upc[i] = normal_cdf(y[i] | theta_i, sqrt(sigma2));
        y_llik[i] = normal_lpdf(y[i] | theta_i, sqrt(sigma2));
    }

    for (i in 1:N) {
        theta_upc[i] = normal_cdf(theta[i] | mu, tau2);
    }
}

```

```{r stan}
suppressPackageStartupMessages(library("rstan"))
suppressPackageStartupMessages(library("goftest"))
suppressPackageStartupMessages(library("loo"))
suppressPackageStartupMessages(library("knitr"))
dt <- read.csv('schools651.csv', header = TRUE)
dt <- dt[, -1] # Remove the rownames
mu_0 <- 7
gamma2_0 <- 5
tau2_0 <- 10
eta_0 <- 2
sigma2_0 <- 15
nu_0 <- 2
N <- 8 # Number of schools
n <- length(dt$school)

data_stan <- list(N = N, n = n, y = dt$times, grp = dt$school, nu_0 = nu_0, sigma2_0 = sigma2_0, eta_0 = eta_0, tau2_0= tau2_0, mu_0 = mu_0, gamma2_0 = gamma2_0)

params <- c("theta", "mu", "sigma2", "tau2", "y_rep", "y_upc", "y_llik", "theta_upc")

n_cores <- parallel::detectCores()   # Count how many cores are available
options(mc.cores = n_cores)          

fit_stan <- stan(model_code = readLines("q1_a_e.stan"),
                 data = data_stan, pars = params,
                 iter = 5000, warmup = 500, thin = 5, chains = n_cores)


sim_stan <- extract(fit_stan) # collect samples from all chains

n_sims <- nrow(sim_stan$theta)
var_diffs <- numeric(n_sims)
for (i in 1:n_sims) {
    var_diffs[[i]] <- diff(range(tapply(sim_stan$y_rep[i, ], dt$school, var)))
}

obs_var_diff <- diff(range(tapply(dt$times, dt$school, var)))
var_diff_p <- mean(var_diffs > obs_var_diff)
```


If the p-value for the range between the smallest and largest variances is large, that would imply that our assumption of shared variance between all the groups is not a reasonable assumption. If the p-value is small, then that would indicate that it is likely a reasonable assumption. A large p-value would imply that we should refit the model using indiviudal group variances, as opposed to one joint population variance.

In this case, the range is `r var_diff_p`. 

### b) 

I propose a test statistic of  the variance of the means: $T(y, \theta) = Var(mean_j(y_{i,j}))$. If this p-value is small, that could be an indication that we actually don't need to the different school means, and could instead use a shared mean across all. 

```{r}
means_var <- numeric(n_sims)
for (i in 1:n_sims) {
    means_var[[i]] <- var(tapply(sim_stan$y_rep[i, ], dt$school, mean))
}

obs_means_var <- var(tapply(dt$times, dt$school, mean))
means_var_p <- mean(means_var > obs_means_var)
```

This has a value of `r means_var_p` meaning that our assumption of different means is probably a good assumption. 

### c) 

##### i) 

Single draw UPC test:
```{r}
indx_upc <- 2000 # pick just one iteration of MCMC
test <- goftest::ad.test(sim_stan$y_upc[indx_upc,], null = "punif") 
p_val <- round(test$p.value, 4)
```

This has a p-value of `r p_val`, which seems to be a good value showing uniformity. 

##### ii) 

Combining p-values of uniformt tests across all posterior samples:

```{r}
upc_pvals_y <- apply(sim_stan$y_upc, 1, function(x) goftest::ad.test(x, null = "punif")$p)
cauchy_y_p <- round(pcauchy(mean(tan(pi * (0.5 - upc_pvals_y))), lower.tail = FALSE), 4)
```

This gives a p-value of `r cauchy_y_p` which would imply that the model is perhaps inqdequately fit. 

### d) 

```{r}
upc_pvals_theta <- apply(sim_stan$theta_upc, 1, function(x) goftest::ad.test(x, null = "punif")$p)
cauchy_theta_p <- round(pcauchy(mean(tan(pi * (0.5 - upc_pvals_theta))), lower.tail = FALSE), 4)
```

Performing the UPC test for all the thetas gives the following p-value: `r cauchy_theta_p`, which implies that our mdoel does capture the behavior of the thetas reasonably well. 

### e)

```{r}
loo1 <- loo::loo(sim_stan$y_llik)
loo1_est <- round(loo1$estimates[1, 1], 3)
loo1_params <- round(loo1$estimates[2, 1], 2)
waic1 <- loo::waic(sim_stan$y_llik)
waic1_est <- round(waic1$estimates[1, 1], 3)
waic1_params <- round(waic1$estimates[2, 1], 2)
```

This model gives a LOO of `r loo1_est` and a WAIC of `r waic1_est`. The models give model parametere estimates of `r loo1_params` (LOO) and `r waic1_params` (WAIC).

These metrics give a lower number of estimated parameters than what we've used. This likely means that some of our parameters are extra, such as maybe some of the schools have similar/shared means, etc. 

### f)

I'm just going to let all the $\sigma^2_j$ be distributed as described in Hoff:

$\frac{1}{\sigma^2_j} \sim Gamma(\frac{\nu}{2}, \frac{\nu \sigma^2_0}{2})$, but I will pick $\nu \sim Exp(\alpha_0)$ where $\alpha_0$ is a parameter we pick, to enforce that it must be positive. Not sure if that does anything funky, but we'll see!

STAN for (f)

```{}
data {
    int<lower=0> N; //Number of Schools
    int<lower=0> n; //Total datapoints
    array[n] real y; // Responses
    array[n] int grp; //Associated group for each response
    real sigma2_0;
    real eta_0;
    real tau2_0;
    real mu_0;
    real gamma2_0;
    real alpha_0;
}

parameters {
    array[N] real theta;
    array[N] real<lower=0> sigma2;
    real mu;
    real<lower=0> nu;
    real<lower=0> tau2;
    
}

transformed parameters {
    array[n] real theta_vec;
    array[n] real sigma2_vec;
    for (i in 1:n) {
        theta_vec[i] = theta[grp[i]];
    }
    for (i in 1:n) {
        sigma2_vec[i] = sigma2[grp[i]];
    }
}

model {
    y ~ normal(theta_vec, sqrt(sigma2_vec));
    theta ~ normal(mu, sqrt(tau2));
    sigma2 ~ inv_gamma(nu/2, nu * sigma2_0/2);
    tau2 ~ inv_gamma(eta_0/2, eta_0 * tau2_0/2);
    mu ~ normal(mu_0, sqrt(gamma2_0));
    nu ~ exponential(alpha_0); 
}

generated quantities {
    vector[n] y_rep;
    vector[n] y_upc; 
    vector[n] y_llik; 
    vector[N] theta_upc; 
    vector[N] sigma2_upc; 

    for (i in 1:n) { 
        real theta_i = theta[grp[i]]; 
        real sigma2_i = sigma2[grp[i]]; 
        y_rep[i] = normal_rng(theta_i, sqrt(sigma2_i)); 
        y_upc[i] = normal_cdf(y[i] | theta_i, sqrt(sigma2_i));
        y_llik[i] = normal_lpdf(y[i] | theta_i, sqrt(sigma2_i));
    }

    for (i in 1:N) {
        theta_upc[i] = normal_cdf(theta[i] | mu, sqrt(tau2));
    }

    for (i in 1:N) {
        sigma2_upc[i] = inv_gamma_cdf(sigma2[i] | nu/2, nu * sigma2_0 / 2);
    }
}


```

```{r}
alpha_0 <- .5
data_stan_2 <- list(N = N, n = n, y = dt$times, grp = dt$school, sigma2_0 = sigma2_0, eta_0 = eta_0, tau2_0= tau2_0, mu_0 = mu_0, gamma2_0 = gamma2_0, alpha_0 = alpha_0)    

params <- c("theta", "mu", "sigma2", "tau2", "y_rep", "y_upc", "y_llik", "theta_upc", "sigma2_upc")

fit_stan_2 <- stan(model_code = readLines("q1_f.stan"),
                 data = data_stan_2, pars = params,
                 iter = 5000, warmup = 500, thin = 5, chains = n_cores)

sim_stan_2 <- extract(fit_stan_2) # collect samples from all chains

n_sims2 <- nrow(sim_stan_2$theta)
var_diffs2 <- numeric(n_sims2)
for (i in 1:n_sims2) {
    var_diffs2[[i]] <- diff(range(tapply(sim_stan_2$y_rep[i, ], dt$school, var)))
}

obs_var_diff_2 <- diff(range(tapply(dt$times, dt$school, var)))
var_diff_p_2 <- mean(var_diffs2 > obs_var_diff_2)

loo2 <- loo::loo(sim_stan_2$y_llik)
loo2_est <- round(loo2$estimates[1, 1], 3)
loo2_params <- round(loo2$estimates[2, 1], 2)
waic2 <- loo::waic(sim_stan_2$y_llik)
waic2_est <- round(waic2$estimates[1, 1], 3)
waic2_params <- round(waic2$estimates[2, 1], 2)

upc_pvals_y_2 <- apply(sim_stan_2$y_upc, 1, function(x) goftest::ad.test(x, null = "punif")$p)
cauchy_y_p_2 <- round(pcauchy(mean(tan(pi * (0.5 - upc_pvals_y_2))), lower.tail = FALSE), 4)

upc_pvals_theta_2 <- apply(sim_stan_2$theta_upc, 1, function(x) goftest::ad.test(x, null = "punif")$p)
cauchy_theta_p_2 <- round(pcauchy(mean(tan(pi * (0.5 - upc_pvals_theta_2))), lower.tail = FALSE), 4)
```

This model gives a LOO of `r loo2_est` and a WAIC of `r waic2_est`. The models give model parametere estimates of `r loo2_params` (LOO) and `r waic2_params` (WAIC).

### g)

Comparing the two models, the second model has slighlty lower LOOIC and WAIC which imply they are somewhat better. However, performing the same UPC measures as with the first model, we get a $y_{UPC}$ p-value of `r cauchy_y_p_2` which is non-signifcant, especially when compared to the first model. The $\theta_{UPC}$ p-value is `r cauchy_theta_p_2` which is approxiamately the same as the first model. Furthermore, when peforming the same Predicitive Performance metric (range of group means), we get a p-value of `r var_diff_p_2` for the second model, which is much higher than the first one. 

Due to these metrics, I think I would still choose to go with the first model, primarily because of the strong evidence from the predictive performance measure which indicates that we are not properly capturing the behavior of the data in the second model.


# 2

$y_i = \alpha + x_i^{\top}\beta + \epsilon_i; \epsilon \sim N(0, 1)l x_{i,j} \sim N(0,1); i = 1, ..., n; j = 1, ..., p; p = 10; \beta = (10, -5, 3, 0, 0, 0, 0, 0, 0, 0)$

```{r}
# Generate sample data
set.seed(1337)
beta_true <- c(10, -5, 3, 0, 0, 0, 0, 0, 0, 0)
alpha_true <- 0
n <- 100
p <- 10
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
y <- alpha_true + X %*% beta_true + rnorm(n, 0, 1)
```

### a)

Assume $\sigma^2$, the variance of $\epsilon_i$ is unknown and include in the model. I will use a conjugate prior of Normal-Inverse-Gamma where $\beta|\sigma^2 \sim N(0, \sigma^2 I); \sigma^2 \sim Inv-Gamma(\alpha_0, \beta_0)$

STAN code:
```{}
data {
    int<lower=0> n; //Total datapoints
    int<lower=0> p; // Total betas
    matrix[n, p] X; // Predictor matrix
    vector[n] y; // Responses
    real alpha_0;
    real beta_0;
    real lambda_0; // scale factor
}

parameters {
    vector[p] beta;
    real alpha;
    real<lower=0> sigma2;
}

model {
    alpha ~ normal(0, 10); // Weekly informative prior
    sigma2 ~ inv_gamma(alpha_0, beta_0);
    beta ~ normal(0, sigma2 * lambda_0);
    y ~ normal(X * beta + alpha, sqrt(sigma2));  // data model
}

generated quantities {
    vector[n] y_llik; 

    for (i in 1:n) { 
        y_llik[i] = normal_lpdf(y[i] | X[i] * beta + alpha, sqrt(sigma2));
    }
}

```

```{r}
alpha_0 <- 2
beta_0 <- 2
mu_0 <- 2
lambda_0 <- 1
data_stan <- list(n = n, p = p, X = X, y = y[,1], alpha_0 = alpha_0, beta_0 = beta_0, lambda_0 = lambda_0)

params <- c("beta", "alpha", "sigma2")

fit_stan_a <- stan(model_code = readLines("q2_a.stan"),
                 data = data_stan, pars = params,
                 iter = 5000, warmup = 500, thin = 5, chains = n_cores)

sim_stan_a <- extract(fit_stan_a) # collect samples from all chains
```

### b)

STAN Code: 

```{}
data {
    int<lower=0> n; //Total datapoints
    int<lower=0> p; // Total betas
    matrix[n, p] X; // Predictor matrix
    vector[n] y; // Responses
    real alpha_0;
    real beta_0;
    real lambda_0; // scale factor
}

parameters {
    vector[p] beta_raw;
    real alpha;
    real<lower=0> sigma2;
    vector<lower=0>[p] lambda;
    real<lower=0> tau;
}

transformed parameters {
  vector[p] beta;
  beta = beta_raw .* lambda * tau * sqrt(sigma2); // Multiply by tau
}

model {
    lambda ~ cauchy(0, 1);
    tau ~ cauchy(0, 1);
    alpha ~ normal(0, 10); // Weekly informative prior
    sigma2 ~ inv_gamma(alpha_0, beta_0);
    beta_raw ~ std_normal();
    y ~ normal(X * beta + alpha, sqrt(sigma2));  // data model
}

```

```{r}
params <- c("beta", "alpha", "sigma2", "lambda", "tau")

fit_stan_b <- stan(model_code = readLines("q2_b.stan"),
                 data = data_stan, pars = params,
                 iter = 5000, warmup = 500, thin = 5, chains = n_cores)

sim_stan_b <- extract(fit_stan_b) # collect samples from all chains
```

### c)

```{r}
sim_stan_a$beta[1,]
a_beta_means <- numeric(p)

a_beta_means <- round(apply(sim_stan_a$beta, 2, mean), 3)
a_beta_quants <- round(apply(sim_stan_a$beta, 2, quantile, probs = c(0.025, 0.975)), 3)
b_beta_means <- round(apply(sim_stan_b$beta, 2, mean), 3)
b_beta_quants <- round(apply(sim_stan_b$beta, 2, quantile, probs = c(0.025, 0.975)), 3)

beta_vals <- c(1:10)
tab <- as.data.frame(cbind(beta_vals, a_beta_means, a_beta_quants[1, ], a_beta_quants[2, ], b_beta_means, b_beta_quants[1, ], b_beta_quants[2, ]))
colnames(tab) <- c('Beta', 'Model 1 Mean', 'Model 1 Lower', 'Model 1 Upper','Model 2 Mean', 'Model 2 Lower', 'Model 2 Upper')
kable(tab)
```

The intervals are consistently smaller on the second model, which implies it's a little bit better. 

### d)

I plot the distributions of $\kappa$ for the first 6 $\kappa$'s (the other 4 are similar to the previous)

```{r}
n_sims <- nrow(sim_stan_b$sigma2)
kappas <- matrix(NA, nrow = n_sims, ncol = p)
for (i in 1:n_sims) {
    kappas[i,] <- 1/(1 + sim_stan_b$lambda[i,]^2 * sim_stan_b$tau[i])
}

par(mfrow = c(2, 3))
hist(kappas[,1])
hist(kappas[,2])
hist(kappas[,3])
hist(kappas[,4])
hist(kappas[,5])
hist(kappas[,6])
par(mfrow = c(1, 1))
```

This shows that yes, this model does appear to correctly include the first 3 parameters, and shrink away the rest. 

### e) 

```{r}
# Generate sample data
set.seed(1337)
beta_true <- c(10, -5, 3, 0, 0, 0, 0, 0, 0, 0)
alpha_true <- 0
n <- 100
p <- 10
sigma <- runif(p)
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
y <- alpha_true + X %*% beta_true + rnorm(n, 0, sigma)
```

STAN code for this given previously

```{r}
alpha_0 <- 2
beta_0 <- 2
mu_0 <- 2
lambda_0 <- 1
data_stan <- list(n = n, p = p, X = X, y = y[,1], alpha_0 = alpha_0, beta_0 = beta_0, lambda_0 = lambda_0)

params <- c("beta", "alpha", "sigma2", "y_llik")

fit_stan_e_1 <- stan(model_code = readLines("q2_a.stan"),
                 data = data_stan, pars = params,
                 iter = 5000, warmup = 500, thin = 5, chains = n_cores)

sim_stan_e_1 <- extract(fit_stan_e_1) # collect samples from all chains

loo1 <- loo::loo(sim_stan_e_1$y_llik)
loo1_est <- round(loo1$estimates[1, 1], 3)
loo1_params <- round(loo1$estimates[2, 1], 2)
waic1 <- loo::waic(sim_stan_e_1$y_llik)
waic1_est <- round(waic1$estimates[1, 1], 3)
waic1_params <- round(waic1$estimates[2, 1], 2)
```

STAN Code: 

```{}
data {
    int<lower=0> n; //Total datapoints
    int<lower=0> p; // Total betas
    matrix[n, p] X; // Predictor matrix
    vector[n] y; // Responses
    real alpha_0;
    real beta_0;
    real lambda_0; // scale factor
}

parameters {
    vector[p] beta;
    real alpha;
    real tau2;
    vector<lower=0>[n] sigma2;
}

model {
    alpha ~ normal(0, 10); // Weekly informative prior
    sigma2 ~ inv_gamma(alpha_0, beta_0);
    beta ~ normal(0, tau2 * lambda_0);
    y ~ normal(X * beta + alpha, sqrt(sigma2));  // data model
}

generated quantities {
    vector[n] y_llik; 

    for (i in 1:n) { 
        y_llik[i] = normal_lpdf(y[i] | X[i] * beta + alpha, sqrt(sigma2[i]));
    }
}

```

```{r}
fit_stan_e_2 <- stan(model_code = readLines("q2_e.stan"),
                 data = data_stan, pars = params,
                 iter = 5000, warmup = 500, thin = 5, chains = n_cores)

sim_stan_e_2 <- extract(fit_stan_e_2) # collect samples from all chains

loo2 <- loo::loo(sim_stan_e_2$y_llik)
loo2_est <- round(loo2$estimates[1, 1], 3)
loo2_params <- round(loo2$estimates[2, 1], 2)
waic2 <- loo::waic(sim_stan_e_2$y_llik)
waic2_est <- round(waic2$estimates[1, 1], 3)
waic2_params <- round(waic2$estimates[2, 1], 2)
```

Model1 gives a LOO of `r loo1_est` and a WAIC of `r waic1_est`. Model2 gives a LOO of `r loo2_est` and a WAIC of `r waic2_est`. Both metrics are lower for the second model (the unequal variances), indicating that it is perhaps a better fit. 


# Acknowledgements

#### Estimated time

I estimate this assignment took me in total around 8 hours to complete.

#### Resources Used

I used GoogleAI to help me understand what the heck to do with Posterior Predictive Checks (ie. how to generate the y_rep, and how to compare that to y_obs), as well as debugging my RStan code, and particularly how to do the various priors for the second question. 